Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-05-23 review/answer_to_review_2.md

> Page 6, Column 2, Lines 46-49. ‘If one believes the sample, then a sudden, radical change from burial in flat graves to hill graves took place in Southern Germany at the beginning of the Middle Bronze Age.’  The sample size here is 8 dates. I appreciate the cautionary ‘if one believes the sample’ but this is a bit of a stretch to say the least, and readers that are numerically less literate might be misled. The same applies to other contexts with extremely small sample sizes and I think adding confidence intervals to those proportions would help.

TODO

> Table 3 & Figure 6. The figure nicely summarises the trend detailed on Table 3, but does not distinguish instances where the correlation were identified as signficant vs those that were not. Indeed in no cases the analysis shows a signficant correlation under the Partial Mantel test, whilst only few chronological blocks are showing signficant correlations. I suggest highlighting this by perhaps changing the the point symbology. It is also worth noting that if anything there is a bias towards type I error given: a) that we are dealing with multiple testing; and b) these analyses do not take into account sample sizes which are often very small. P-values are not the core issue here, but rather the fact that some way to at least visually convey this uncertainty is much needed.

I added a visualisation of the p-values to figure 6 now. I understand your concerns, but I do not know how I could modify the figure with the available data to convey all the underlying uncertainty. A permutation test as suggested above may be the way forward here. I added a sentences on page 5 to explain the figure with these problems in mind.

### 2019-05-26 article/05_Conclusion.Rmd
some important changes to the conclusion

*The main trends in the distribution of burial rites in Bronze Age Europe can be detected in bulk radiocarbon data* and indeed *they allow long-term observations concerning cultural distance*. Radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves -- a small fraction of maybe hundreds of thousands burials that can be estimated for the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] -- have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this interpretation must be treated with caution, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.

Future research is required to untangle these interactions -- of course on the basis of more precise data. This endeavour will not end with the simple description of social learning cascades, but with the analysis of the initial reasons for diversification. From the point of view of Cultural complexity -- the question, which factors accelerate or decelerate the variability of human behaviour [@henrichDemographyCulturalEvolution2004; @mesoudiWhatCumulativeCultural2018; @vaesenPopulationSizeDoes2016] -- the long term dominance fluctuations of one burial rite over the other are remarkable. Its standard explanation models (among others population density and environmental risk) can hardly be applied to stylistic variables.

Beyond these intrinsic results, it is also important to ask what this paper means for the study of human behaviour in general. Two possible answers:

>>>>

*The main trends in the distribution of burial rites in Bronze Age Europe can be detected in bulk radiocarbon data* (Figure \ref{development_burial_type} B and \ref{development_burial_construction} B) and indeed *they allow long-term observations concerning cultural distance* (Figure \ref{sed_region_matrix}, Table \ref{tab:dacta} and Table \ref{tab:dactb} and Figure \ref{correlationtimeseries} B), although with the low amount of data currently available, *most findings should be taken with great caution* (Table \ref{tab:gprcrosstab}, Figure \ref{development_burial_type} A and \ref{development_burial_construction} A, Figure \ref{correlationtimeseries} A). In this paper sample size is effectively being ignored once proportions of burial rites are computed.

Even small amounts of radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves -- a small fraction of maybe hundreds of thousands burials that can be estimated for the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] -- have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this interpretation must be treated with caution, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.

From the point of view of Cultural complexity -- the question, which factors accelerate or decelerate the variability of human behaviour [@henrichDemographyCulturalEvolution2004; @mesoudiWhatCumulativeCultural2018; @vaesenPopulationSizeDoes2016] -- the long term dominance fluctuations of one burial rite over the other are remarkable. Its standard explanation models (among others population density and environmental risk) can hardly be applied to stylistic variables.

Future research is required to untangle these interactions -- of course on the basis of more and more precise data. In this future work the methodology should be refined and extended in several ways. Some important ideas:

1. Replacing time slices in the correlation analysis by a moving window approach to get a contious interaction proxy time series.

2. Applying a model based analysis of the idea proportion adoption curve time series and their regional differences to get a better understanding of mutual influence, earliest uptake and laggards.

3. Comparing the real world correlation time series with ones constructed from simulated, randomly permutating observations to test the significance of seemingly positive or negative correlation.

4. Abolishing the separation into artificial regions for the cultural distance and correlation analysis in favour of an approach based on individual observations [@loogEstimatingMobilityUsing2017].

Finally it is also important to ask what this paper means for the study of human behaviour in general. Two possible answers:
